Biodiversity changes are linked to changes in food-web structure and redundancy in riverine ecosystems

### 2025-02-03 paper/manuscript.Rmd
Manuscript update before Maud & Colin comments

---
title: "Temporal dynamics of food-web structure in response to biodiversity changes in riverine ecosystems"
author: "Alain Danet (a,b), Loubna El Madouri (a,c), Maud Mouchet (a), Elisa Thébault (d) and Colin Fontaine (a)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(b) Department of Biosciences, Alfred Denny Building, University of Sheffield,
Western Bank, Sheffield S10 2TN, United Kingdom

@@

- Keywords: food-web, biodiversity changes, assembly rules, space-for-time

# Abstract (250 / 250 words max)

While there are widespread evidence of biodiversity decline, little is known how
food-web disassemble in response. Such knowledge is crucial as food-web
structure determine ecosystem functioning and stability to perturbations.
However, assessing how changes in biodiversity affect changes in food-web
structure require the availability of food-web timeseries and to account for the
differential effects of different biodiversity facets on food-web structure.
Here, we used a standardized monitoring of riverine fish communities over the
last decades (1995-2018) in France, from which we inferred the food-web
structure for each sampling events using diet data and allometric scaling rules.
We estimated the temporal trends in community biomass, species richness, and
food-web structure using hierarchical Bayesian models. We assessed the effects
of community biomass and species richness temporal changes on temporal changes
in food-web structure: connectance, average trophic level and redundancy. We
further assess the generality and robustness of our findings by replicating the
analysis over space and using theoretical simulations with a bioenergetic model.
Our results suggest that food-webs that experienced a decline in community
biomass experienced conjointly a loss of species at upper trophic levels and
conversely. While species richness temporal trends had less clear effects on
food-web structure, mainly due through its indirect effects on community biomass
temporal trends. Both community biomass and species richness declines were
associated with a decrease in redundancy, raising concerns about robustness of
such communities to further perturbations. Our study highlights how biodiversity
changes affect food-web structure, raising concerns about consequences on
ecosystem functioning and robustness.

# Significance (109 / 120 words max)

Biodiversity temporal changes can lead to simultaneous changes in food-web
structure, which is turn can have consequences on ecosystem functioning and
stability. We provide empirical evidence that temporal changes in community
biomass and species richness are linked to changes contrasted assembly and
disassembly processes in food-webs in natural settings. We then show that loss
in both community biomass and species richness are related to a loss of
redundancy in the food-webs, which raise concerns about the robustness of such
communities to perturbations. We further added spatial and theoretical evidence
suggesting that our results are quite robust and open to assess how biodiversity
changes might affect ecosystem functioning and robustness.

@@

Biodiversity is experiencing massive declines due to continuously increasing
human pressures, which is expected to have striking consequences on ecosystem
functioning and fragility. It is increasing recognised that biodiversity
changes are widely heterogeneous spatially [@mcgill_fifteen_2015;
@blowes_geography_2019], such as species richness declines and increases can be
related to spatial variations in human pressures [@cardinale_is_2018;
@danet_past_2024]. Biodiversity changes result in altered community structure
[@dornelas_assemblage_2014; @hillebrand_biodiversity_2018; @danet_past_2024],
but so far, one key component of biodiversity, species interactions, has been
mostly ignored by studies assessing temporal changes in biodiversity, despite
the importance of biotic interactions for species coexistence, ecosystem
functioning and stability [@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017; @danet_species_2021]. A better understanding of
biodiversity changes will then need to account how biodiversity changes alter
the structure of ecological networks.

Yet, both species richness and food-web structure have a significant impact on
coexistence [@chesson_interaction_2008], ecosystem functioning
[@thebault_food-web_2003; @wang_biodiversity_2018], stability
[@danet_species_2021; @eschenbrenner_diversity_2023], and ecosystem robustness
to perturbations [@dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @allesina_functional_2009]. Many predator species play a
keystone role by protecting preys from competitive exclusion
[@heithaus_predicting_2008; @johnson_rarity_2007; @estes_trophic_2011], thereby
maintaining species richness. Food-web with higher trophic levels also increase
ecosystem productivity [@thebault_food-web_2003; @wang_biodiversity_2018] and
community temporal stability [@danet_species_2021;
@eschenbrenner_diversity_2023]. Finally, the number and distribution of trophic
links in food-webs determine the robustness of communities to perturbations
[@dunne_network_2002; @allesina_functional_2009; @bodini_using_2009;
@allesina_functional_2009]. Accounting for how food-web assemble and disassemble
according to biodiversity changes is then crucial to evaluate the consequences
of biodiversity changes on ecosystem functioning and stability.

However, assessing the effects of biodiversity changes on food-web structure
requires the availability of multiple temporal timeseries of biodiversity and
food-webs structure, to disentangle the direct and indirect effects of
biodiversity changes on food-web structure, and to account for theories about
the multiple constraints on food-web assembly. We propose to monitor three
structural properties of food-webs and two common monitored biodiversity facets
(i.e. community biomass and species) that are grounded in food-web assembly,
ecosystem functioning, and stability research areas: connectance (i.e. degree of
generalism), weighted average trophic level (i.e. vertical location of biomass),
and redundancy (i.e. the proportion of redundant links). There are pieces of
evidence to connect those food-web structural metrics to changes in ecosystem
functioning and robustness that we review before to present the temporal
timeseries that allows to assess them.

Food-web assembly theory provides predictions on how food-web structure should
change according to changes in community biomass and species richness. The
energetic paradigm and the trophic theory of biogeography
[@lindeman_trophic-dynamic_1942; @gravel_trophic_2011] stand that the presence
of top predators is conditional on the presence of preys in number enough that
it can provide enough energy from the bottom to the top to sustain high trophic
level species. In turn, it can that the presence of generalist predators, which is
positively associated to average trophic level and connectance, will be linked
more to community biomass changes than to species richness. In turn, connectance
and average trophic level are commonly reported to decrease with as species
richness increases [@dunne_network_2006; @danet_species_2021] because predators
are limited in the number of species they can consume and conversely preys are
limited in the number of predator they can sustain. Food-web structure has also
been linked to the robustness to species extinctions [@dunne_network_2002;
@allesina_who_2004; @bodini_using_2009]. Historically having being linked to
connectance, i.e. the number of links, more recent studies have proposed to
separate links that are functional or redundant [@allesina_who_2004;
@allesina_functional_2009; @bodini_using_2009]. The idea is that the loss of
functional links will trigger secondary extinctions because it suppresses
obligatory pathways from resources to consumers while the loss of redundant
links do not remove such obligatory pathways [@allesina_who_2004]. It means that
redundant trophic links enhance ecosystem robustness as they can be safely
removed without triggering secondary extinctions. Thereby we expected that
higher species richness results in higher redundancy in food-webs, meaning that
they create more independent pathways connecting basal producers to higher
trophic level. Finally, biodiversity and ecosystem functioning theory suggest
that in context of significant niche complementarity, community biomass should
be positively related to species richness [@wang_biodiversity_2018;
@schneider_animal_2016], such as it is necessary to account for the indirect
effects of species richness on food-web structure through community biomass.

Here, we report on a study that combine empirical assessment of food-web
structure and biodiversity temporal changes in natural settings, extinction
simulations in empirical food-webs, and theoretical simulations using
bioenergetic model. The availability of food-web timeseries used to be scarce
[@kortsch_disentangling_2021], but the increasing availability of biodiversity
timeseries and the development of methods to reconstruct food-webs
[@bonnaffe_comparison_2021; @gravel_inferring_2013; @danet_species_2021] makes
possible to overcome this limitation. We used a standardized monitoring of fish
riverine communities on the period 1995-2018 over France, documenting the
individual size of fishes in `r nb_st` sites during ten years or more (Table S1). We
reconstructed around `r nb_op` food-webs based on allometric relationships and
food diet data, and then computed connectance, and average trophic level
weighted by trophic species biomass as well as community biomass and species
richness (Fig. \@ref(fig:matmet)). This dataset and method has already been used
in previous studies [@danet_species_2021; @bonnaffe_comparison_2021;
@bonnaffe_interaction_2024]. Using hierarchical Bayesian models, we estimated
temporal trends in food-web structure, community biomass and richness at each
site [@danet_past_2024]. We then used structural equation models (SEM) to
assess the effects of community biomass and species richness temporal changes on
changes in food-web structure, while accounting for the effects of species
richness on community biomass. We then assessed if those effects are holding in
space, i.e. if the space-for-time substitution hypothesis hold
[@blois_space_2013; @damgaard_critique_2019; @adler_contrasting_2007]. We did so
by considering median values of food-web structure and community structure over
the timeseries at each site (Fig. \@ref(fig:matmet)). Finally, we ran
simulations using the bioenergetic model to assess if our results hold beyond
our empirical fish food-webs.

```{r}
fig_mat_cap <- paste0("Using sites of stream fishes in French streams that
  have been monitored at least 10 years, 1) we inferred the food-web at each site
and each sampling event based on ontogenic food diet information for each
species and predator-prey body size ratio. 2) We computed the temporal trends
and the median of community biomass, species richness, connectance, average
trophic level and evenness of biomass at each site. 3) We then assessed the
relationships among food-web structure descriptors, (a) in time and (b) across space.")
```

@@

Using a Bayesian hierarchical model to jointly estimate the temporal trends of
food-web structure at all sites, we found weak evidence of an overall decrease
in community biomass (CI80%:`r gbm` by decade; median [Credible Interval], Fig.
S1) but no evidence of an overall temporal increase in species richness
(CI80%: `r gr` by decade) across the `r nb_st` sites. We did not find evidence
of average temporal trends in connectance and average trophic level (CI80%:
`r gct` and `r gt` respectively), but an increase in redundancy (measured by the
proportion of redundant links, CI80%: `r gl`). However, we found a great
heterogeneity in the temporal trends estimated at the site level (community
biomass: from `r grbmn`% to `r grbmx`% per decade, species richness: `r grrhn`
to +`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig. S1) which was
well captured by the hierarchical model with conditional R2 going from
`r r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std` for
species richness (Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of the food-web
structure. The path widths are proportional to the standardized slope
coefficients. The standardized slope coefficients and their associated 95%
confidence intervals obtained by bootstrapping are displayed along the path. The
marginal R-squared of the linear models are displayed. (b to f) Direct bivariate
relationships derived from the structural equation model. The distribution of
the site temporal trends are displayed in Fig. S1. All the effects are reported
in Table S3."
```

@@

Using structural equation models to disentangle the direct and indirect effects
of temporal trends in fish species richness and community biomass on temporal
trends in food-web structure, we found a strong positive effect of the temporal
trends in species richness on community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3), meaning that both temporal trends had the same direction in average. In
other word, food-webs that have experienced a decrease in species richness over
time have decreased in community biomass, and conversely, those that increased
in species richness have increased in community biomass.
Temporal trends in community biomass had in turn a positive effect on connectance,
average trophic level, and redundancy (resp. $r_\delta =$ `r rdtps$ct$bm`,
`r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig. \@ref(fig:sem)b, c, d), meaning that
food-web whose community biomass decreased over time have decreased in
connectance, average trophic level and redundancy, and
conversely, those that have experienced an increase in biomass have increased in
connectance, average trophic level and redundancy. We further found that the temporal
trends of community biomass were positively associated to the temporal trends of
both number and proportion of piscivorous species present (Fig. S2a), meaning
biomass trends are linked to the presence of high trophic level trophic
species.

We further tested if we find the same patterns using spatial variations in
food-web structure rather than temporal changes over time, i.e. if the
space-for-time hypothesis hold. To do so, we used the same dataset and same
structural equation model but with the median food-web structure per site
instead of the temporal trends. We additionally performed theoretical
simulations with a bioenergetic food-web model to further get theoretical
expectations about how food-web structure changes according to community biomass and
species richness variations (see Methods). As in temporal data, species richness
had a positive effect on community biomass both in empirical data and in
theoretical simulations (resp. $r_\delta =$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig.
\@ref(fig:std) and S3). In both empirical data and theoretical simulations, community
biomass had a positive effect average trophic level (resp.
$r_\delta =$ `r rdsp$tlvl$bm` and `r rdm$tlvl$bm`), and redundancy (resp.
$r_\delta =$ `r rdsp$red$bm` and `r rdm$red$bm`), as in temporal analysis.
We found weak evidence of a positive effect of spatial variation in community
biomass on connectance in empirical data ($r_\delta =$ `r rdsp$ct$bm`) but
strong evidence in theoretical simulation ($r_\delta =$ `r rdm$ct$bm`, Fig.
\@ref(fig:std)).

The effects of species richness were more complex, because species richness had
indirect effects on food-web structure through community biomass. Temporal
trends in species richness had a direct negative effect on connectance
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive
indirect effect through community biomass ($r_\delta =$ `r ridtps$ct$r`), so
that species richness temporal trends had a negative effect in total on temporal in
connectance ($r_\delta =$ `r rttps$ct$r`, Fig. \@ref(fig:std)).
In contrast, temporal trends in species had no direct effects on the temporal
trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), a positive indirect effect through community
biomass ($r_\delta =$ `r ridtps$tlvl$r`), which adds up to a positive total
effect of species richness temporal trends on average trophic level ($r_\delta
=$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)). Species richness temporal trends had
direct and indirect positive effect on the temporal trends in redundancy (resp.
$r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong
positive effect ($r_\delta =$ `r rttps$red$r`). Furthermore, we found that
temporal trends of the proportion of piscivorous species was negatively
associated with the temporal trends of species richness (Fig. S2), meaning that
communities that increase in richness over time have an increase in the
proportion of non-piscivorous species over time and vice versa. It means
food-webs that have experienced declines in species richness over the last
decades have decreased in average trophic level and redundancy but increase in
connectance, and conversely, those that experienced increases in species
richness have increased in average trophic level, redundancy, but decreased in
connectance.

The effects of spatial variations in species richness on food-web structure
mostly agreed with the ones found with temporal trends, expect for average
trophic level. We found a direct negative effect of species richness on
connectance both using data and model simulations
(resp. $r_\delta =$ `r rdsp$ct$r` and `r rdm$ct$r`, Fig. \@ref(fig:std)), on
average trophic level (resp. $r_\delta =$ `r rdsp$tlvl$r` and `r rdm$tlvl$r`),
but a direct positive effect on redundancy (resp. $r_\delta =$ `r rdsp$red$r`
and `r rdm$red$r`). Species richness had indirect positive effect of species
richness through community biomass, which added up to strong total negative
effects on connectance in empirical data and simulations (resp. $r_\delta =$
`r rtsp$ct$r` and `r rtm$ct$r` ) and positive on redundancy (resp. $r_\delta =$
`r rtsp$red$r` and `r rtm$red$r`). We found that species richness had a total
positive effect on average trophic level in bioenergetic model, in agreement
with temporal trends (resp. $r_\delta =$ `r rtm$tlvl$r`), but a total negative
effect in spatial variations (resp. $r_\delta =$ `r rtsp$tlvl$r`).

We performed extinction simulations scenario on the metaweb, the food-web
including all the species and nodes found in the empirical data, to better
understand how species richness change affect food-web structure. When species
are removed in a sequential order beginning by the highest trophic levels,
connectance and average trophic level decreased as we removed species (red
lines, Fig. S4a). At the opposite, removing first low trophic level species first
increase connectance and average trophic level (green lines, Fig. S4a). In turn,
random species extinction sequence had little effect on food-web structure
(black lines, Fig. S4a).

```{r}
std_cap <- paste0(
  "Direct and total effects of community biomass and species richness on
  food-web structure. Standardized coefficients ($r_\\delta$) derived from the
  temporal, spatial, and bioenergetic model Structural Equation Models. Points
  and lines display the estimated effects and their confidence intervals at
  95\\%. Values are displayed in Table S3, R-squared in Table S4, and the
  visualisation of the bivariate relationships for the spatial and bioenergetic
  model SEMs are displayed in Figure S3.")

```

```{r std, fig.cap=std_cap, out.width = "100%"}
tar_load(p_semeff_tot_ok_file)
knitr::include_graphics(p_semeff_tot_ok_file)
```

>>>>

---
title: "Biodiversity changes are linked to changes in food-web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(b) School of Biosciences, Alfred Denny Building, University of Sheffield,
Western Bank, Sheffield S10 2TN, United Kingdom

@@

- Keywords: food web structure, biodiversity changes, redundancy, biodiversity &
  ecosystem functioning

# Abstract (152 / 250 words max)

Changes in biodiversity can fundamentally alter ecological communities, yet how
these changes affect food web structure remains poorly understood. Here, we
analyze temporal trends in over 400 riverine fish communities across France
spanning 1995-2018, examining relationships between species richness, community
biomass, and three key metrics of food web structure: connectance, weighted
average trophic level, and trophic pathway redundancy. Using structural equation
modeling, theoretical simulations, and robustness analyses, we demonstrate that declining
community biomass leads to loss of top trophic levels and decreased food web
connectance, while species richness negatively correlates with connectance but
positively affects trophic redundancy. Notably, both biomass and species
richness positively influence the proportion of redundant trophic links,
suggesting that biodiversity loss reduces community robustness to future
perturbations. Our findings, consistent across temporal, spatial, and
theoretical approaches, reveal how biodiversity change restructures food webs
and highlights the utility of food web metrics as indicators of ecosystem
function and fragility.

# Significance (124 / 120 words max)

Understanding how biodiversity loss affects ecological communities is crucial
for predicting and mitigating human impacts on ecosystems. This study
provides comprehensive evidence linking biodiversity changes to food web
structure using an extensive dataset of riverine fish communities. We
demonstrate that changes in community biomass and species richness
systematically alter food web architecture, affecting trophic levels,
connectance, and the redundancy of feeding relationships. Our findings unite
multiple ecological theories, showing how energetic constraints and
biodiversity-ecosystem functioning mechanisms jointly shape food web assembly
and disassembly. Importantly, we show that declining biodiversity reduces the
redundancy of trophic pathways, potentially making communities more vulnerable
to future perturbations. These results suggest that food web structure metrics
could serve as valuable indicators for monitoring ecosystem health and
evaluating conservation strategies.

@@

Ecological communities are experiencing massive changes due to human pressures,
which are expected to have striking consequences on ecosystem functioning and
fragility. It is increasingly recognized that such changes across space are
associated with spatial variations in human pressures [@mcgill_fifteen_2015;
@blowes_geography_2019], such as freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024], and declining
biomass of marine fishes linked to increased fishing pressures on predatory
fishes [@christensen_century_2014]. These documented changes in diversity and
biomass result in altered community structure and composition
[@dornelas_assemblage_2014; @hillebrand_biodiversity_2018; @danet_past_2024].

The structure of ecological networks, which describe the number, distribution
and strength of interactions among species, is important because it, along with
species richness per se, underpins species coexistence, ecosystem functioning
and the stability of ecological dynamics and ecosystem function
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024].
Theory and empirical data have shown that species richness and food-web
structure impact on coexistence [@chesson_interaction_2008], ecosystem
functioning [@thebault_food-web_2003; @wang_biodiversity_2018], stability
[@danet_species_2021; @danet_response_2024; @eschenbrenner_diversity_2023], and
ecosystem robustness to perturbations [@dunne_network_2002;
@allesina_functional_2009; @bodini_using_2009; @allesina_functional_2009].
Keystone predator species regulating dominant competitors can limit competitive
exclusion [@heithaus_predicting_2008; @johnson_rarity_2007;
@estes_trophic_2011], and thereby maintain species richness. Food-webs with
higher trophic levels are linked to elevated  ecosystem productivity
[@thebault_food-web_2003; @wang_biodiversity_2018] and community temporal
stability [@danet_species_2021; @eschenbrenner_diversity_2023;
@danet_response_2024]. Finally, the number and distribution of trophic links in
food-webs determine the robustness of communities to perturbations
[@dunne_network_2002; @allesina_functional_2009; @bodini_using_2009;
@allesina_functional_2009]. A better knowledge of the consequences of
biodiversity change on the structure of ecological networks is thus necessary to
quantify and mitigate anthropogenic driven changes to biodiversity.

There are three key components to the mapping between biodiversity changes and
food-web structure. First, the energetic paradigm underpinning the trophic
theory of biogeography [@lindeman_trophic-dynamic_1942; @gravel_trophic_2011]
posits that the presence of top consumers requires resources in enough numbers
at lower trophic levels derived from bottom up productivity. From this, it can
be further hypothesized that the presence of generalist top consumers will be
strongly conditioned by the amount of community biomass.

Second, connectance (the proportion of realized trophic links), is commonly
reported to decrease as species richness increases [@oconnor_general_2017;
@dunne_network_2006], because consumers are limited in the number of resources
they can consume. Connectance has also been linked to the robustness to species
extinctions [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009]. More
in detail, it is the prevalence of redundant (versus functional) links in food
webs [@allesina_who_2004; @bodini_using_2009; @allesina_functional_2009], i.e.
the redundancy of trophic pathways, that predicts ecosystem robustness as they
can be safely removed without triggering secondary extinctions. We thus predict
that higher species richness covaries with higher redundancy of trophic pathways
in food-webs [@allesina_functional_2009]. Together, these relationships mean
that biodiversity changes are likely to be associated with changes in trophic
links distribution and then possibly to changes in robustness to perturbations,
such as increases in species richness should increase the number of redundant
trophic pathways to top predators and then increase community robustness.

Finally, biodiversity and ecosystem functioning theory suggest specifically that
community biomass should be positively related to species richness
[@loreau_biodiversity_2001]. This arises specifically because of niche
complementarity where species exploit different resources. In the particular
case of food webs, theoretical studies show that the presence of higher trophic
level species decreases niche overlap among producers which then increase
productivity and community biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023].

Here, we report on an analysis of riverine fish community times series to
document the ecological consequences of biodiversity changes on food web
structure. We specifically aim to reveal the relationship between community
biomass and species richness and three key metrics of food web structure:
connectance, weighted average trophic level, and the proportion of redundant
trophic links (hereafter “redundancy”). To do this, we implemented a sequence of
analyses that allows us to enumerate these relationships as covariation among
temporal trends. First, we estimated `r nb_op` food web structure for `r nb_st`
fish communities with 10+ years of community dynamics spanning 1995-2018 over
France (Fig. \@ref(fig:matmet)a-c, Table S1).  Second, using these food webs, we
estimated the temporal trends in each of species richness, biomass, connectance,
trophic structure and redundancy. Having estimated each temporal trend, we next
constructed an SEM to make specific inference about the relative effects of
community biomass and species richness  temporal trends on the temporal trends
three metrics of food web structure, and the effect of species richness on
community biomass (Fig. \@ref(fig:matmet)d). We also implemented a robustness analysis on the
meta-food web (i.e. the food web grouping all trophic species found in the
dataset and their inferred interactions) to identify species extinction
scenarios that might explain the variation in food web structure that we see in
the timeseries. Finally, we performed three further analysis to evaluate the
robustness and the generality of our results. We replicated our temporal
analysis with a space-for-time approach (Fig. \@ref(fig:matmet)c & e) using median metrics in
each site instead of the temporal trends; we created a simulation of complex
communities using a bioenergetic food web model to assess the generality of our
results in complex communities that allow relaxation of some of the constraints
defined by our fish biodiversity data; and we performed a three-scenario,
topological secondary extinction analysis on the full empirical metaweb to
evaluate the role of top predators on food-web structure.

```{r}
fig_mat_cap <- paste0("Using ", nb_st," sites of stream fishes in (a) French streams that
  have been monitored at least 10 years, (b) we reconstructed the food-web at each
  site and each sampling event by building a metaweb based on ontogenic food diet information for each
  species and predator-prey body size ratio. (c) We computed the temporal trends
  and the median of community biomass, species richness, connectance, average
  trophic level, and redundancy at each site. We then assessed the relationships
  among food web structure descriptors, (d) in time and (e) across
  space. We additionally performed extinction simulations from the metaweb and
  insilico experiments using a bioenergetic food web model to gain more insights
  and assess the generality of our results (see Methods).")
```

@@

## Estimating temporal trends

Our food web reconstruction delivered `r nb_op` estimates of species richness,
biomass, connectance, weighted average trophic level, and the proportion of
redundant trophic links (hereafter "redundancy", see Appendix B) from our 403
fish communities with 10+ years of population dynamics spanning 1995-2018 over
France (Fig. 1a-c). We implemented Bayesian hierarchical models to jointly
estimate the temporal trends in these metrics. We found a great heterogeneity in
the temporal trends at the site level (community biomass: from `r grbmn`% to
`r grbmx`% per decade, species richness: `r grrhn` to +`r grrhx`%, average
trophic level: `r grtln` to +`r grtlx`, connectance: `r grctn` to +`r grctx`,
redundancy: `r grln` to +`r grlx`, Fig. S1) which was effectively captured by
the hierarchical model (with conditional R2 between `r r2cond$log_bm_std` for
community biomass to `r r2cond$log_rich_std` for species richness, Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of the
food-web structure. The path widths are proportional to the standardized slope
coefficients. The standardized slope coefficients and their associated 95%
  confidence intervals obtained by bootstrapping are displayed along the path.
R2 displays the marginal R-squared of the linear models. (b to h) Direct
bivariate relationships derived from the structural equation model. Points
display the temporal trends at each site estimated with a hierarchical Bayesian
model, lines display the linear relationship estimated with the SEM. The
distribution of the site temporal trends are displayed in Fig. S1. All the
effects are reported in Table S3. Average trophic level is weighted by the
biomass of the trophic species. Redundancy is the proportion of redundant
links."
```

@@

## Association between biodiversity and food-web temporal trends

These temporal trend estimations at the site level were the input to our SEM,
allowing us to evaluate the direct and indirect effect of species richness and
community biomass changes on food web structure changes and between species
richness and community biomass. We detected a strong positive association
between community biomass and in species richness ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3). This indicates that communities with decreasing species richness
over time experienced a concomitant decrease in community biomass and
vice-versa.

Our SEM also predicts a positive effect of community biomass on connectance,
average trophic level, and redundancy (resp. $r_\delta =$ `r rdtps$ct$bm`,
`r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig. \@ref(fig:sem)b, c, d). Thus, when communities
experienced a decrease in biomass, they also experimented a concomitant decrease
in connectance, average trophic level and redundancy, and vice versa. We further
found that the temporal trends in community biomass were positively associated
with both the number and the proportion of piscivorous species (Fig. S2a),
suggesting that the presence of high trophic level trophic species has a
disproportionate role community biomass.

The effects of species richness were more complex, because species richness in
this model has indirect effects on food-web structure through community biomass.
Species richness had a direct negative effect on connectance ($r_\delta =$
`r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive indirect effect through
community biomass ($r_\delta =$ `r ridtps$ct$r`), so that species richness had a
negative effect in total on connectance ($r_\delta =$ `r rttps$ct$r`, Fig.
\@ref(fig:std)a). In contrast, species richness had no direct effects on in
average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)g), a
positive indirect effect through community biomass ($r_\delta =$
`r ridtps$tlvl$r`), which adds up to a positive total effect of species richness
temporal trends on average trophic level ($r_\delta =$ `r rttps$tlvl$r`, Fig.
\@ref(fig:std)a). Species richness temporal trends had direct and indirect
positive effect on the temporal trends in redundancy (resp. $r_\delta =$
`r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong positive effect
($r_\delta =$ `r rttps$red$r`).

Furthermore, we found that temporal trends of the proportion of piscivorous
species was negatively associated with the temporal trends of species richness
(Fig. S2), meaning that communities that increase in richness over time have an
increase in the proportion of non-piscivorous species over time and vice versa.
It means food-webs that have experienced declines in species richness over the
last decades have decreased in average trophic level and redundancy but increase
in connectance, and conversely, those that experienced increases in species
richness have increased in average trophic level, redundancy, but decreased in
connectance.

## Theoretical simulations and space-for-time analysis (mostly) agree with the temporal analysis

We further assessed the consistency of the reported associations among temporal
trends by implementing two additional analyses. We first tested if the
associations among the temporal trends within sites was holding across sites,
i.e. using spatial variations in food-web structure rather than temporal changes
over time (space-for-time hypothesis). Second, we performed theoretical
simulations with a bioenergetic food-web model to cross-reference our site and
species specific empirical data on riverine fish with a more generalized
representation of food web dynamics known to represent accurate relationships
among our focal variables [@brose_allometric_2006; @eschenbrenner_diversity_2023;
@danet_response_2024]. With those additional empirical and theoretical data, we
assessed additional SEM models with the same structure as in Section 4.2.

As in the temporal analysis, species richness had a positive effect on community
biomass both in empirical data and in theoretical simulations (resp. $r_\delta
=$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig. \@ref(fig:std)a and S3). In both
empirical data and theoretical simulations, community biomass had a positive
effect on average trophic level (resp. $r_\delta =$ `r rdsp$tlvl$bm` and
`r rdm$tlvl$bm`), and redundancy (resp. $r_\delta =$ `r rdsp$red$bm` and
`r rdm$red$bm`), as in temporal analysis. We found weak evidence of a positive
effect of spatial variation in community biomass on connectance in empirical
data ($r_\delta =$ `r rdsp$ct$bm`) but strong evidence in theoretical
simulations ($r_\delta =$ `r rdm$ct$bm`, Fig. \@ref(fig:std)a).

The effects of spatial variations in species richness on food-web structure
mostly aligned with those from the temporal trend analysis, except for average
trophic level. We found a direct negative effect of species richness on
connectance both using data and model simulations (resp. $r_\delta =$
`r rdsp$ct$r` and `r rdm$ct$r`, Fig. \@ref(fig:std)a), on average trophic level
(resp. $r_\delta =$ `r rdsp$tlvl$r` and `r rdm$tlvl$r`), but a direct positive
effect on redundancy (resp. $r_\delta =$ `r rdsp$red$r` and `r rdm$red$r`).
Species richness had an indirect positive effect of species richness through
community biomass, which added up to strong total negative effects on
connectance in empirical data and simulations (resp. $r_\delta =$
`r rtsp$ct$r` and `r rtm$ct$r`) and positive on redundancy (resp. $r_\delta =$
`r rtsp$red$r` and `r rtm$red$r`). We found that
species richness had a total positive effect on average trophic level in
bioenergetic model, in agreement with temporal trends (resp. $r_\delta =$ `r rtm$tlvl$r`),
but a total negative effect in spatial variations (resp. $r_\delta =$ `r rtsp$tlvl$r`).

## Robustness of food web structure to species extinctions

We implemented three secondary extinction simulation scenarios on the empirical
metaweb defined as a food-web containing all the trophic species found across
all sites in the empirical data. This was to access a second line of inference
about whether top species were drivers of biomass and structure patterns (see
section 4.2). When species are removed in a sequential order beginning by the
highest trophic levels, connectance, average trophic level, and redundancy
decreased (purple lines, Fig. \@ref(fig:std)b). On the other hand, removing low trophic level
species first increases connectance and average trophic level, but do not affect
redundancy (yellow lines, Fig. \@ref(fig:std)b). In turn, random species extinction sequence
had little effect on food-web structure (black line, Fig. \@ref(fig:std)b). These results, in
tandem with our empirical analyses above, suggest that top consumers play a
disproportionate role in driving the changes in food web structure that we
report.

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food-web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S3, R-squared in Table S4, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Figure S3. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S4).")
```

```{r std, fig.cap=std_cap, out.width = "100%"}
tar_load(p_sem_tot_ext_sp_file)
knitr::include_graphics(p_sem_tot_ext_sp_file[[1]])
```
